Online Sequential Monte Carlo smoother for partially observed diffusion processes

### 2017-09-14 TexSubmission/answer_referees.tex

The importance of these quantities is now clearly stated after equation (2). We added the usual applications to the computation of the Fisher score and the intermediate quantity of the EM algorithm. We also highlighted that it was crucial for online inference of latent data models which supports the use of the proposed online smoother.

\item {\em General practical strategies to specify $\vartheta_k$, $p_k$, $\hat{\sigma}_+^k$ in the SDE context should be
discussed.}

\vspace{.3cm}

Choices that can be made for  $\vartheta_k$ and $p_{k-1}$ are discussed after (4) and the definition of $\phi_k^N$. The specific approximation of the fully adapted particle filter is also illustrated in the numerical section. The $\widehat{\sigma}_+^k$ is an upperbound that must be computed depending on which of the assumptions (A1)-(A3) is satisfied, it is not, in this sens a user's choice. However, as discussed at the end of page 8, it can be better, even if (A1) is satistified, to choose the bound defined by (A3). The first one would (in general) require numerical optimization and provide a general bound (potentially far from the one obtained with the set of particles). The other would require $N^2$ computations ($N$ being the number of particles), but provides the optimal bound for the acceptance reject algorithm of Lemma 1. In addition to the existing paragraph, this is more explicitely discussed now in the numerical section, specially for the SINE model.

@@

This is now mentioned at the end of Section~3.

\item {\em Concerning the paragraph starting in page 8 and ending in page 9. Can the authors
provide some analytical result about when that strategy is better as a function of the
rejection sampling global accep. prob. and the cost to sample (layered) Brownian
bridges?}

\vspace{.3cm}

This important problem is model dependent. Indeed, if assumption $(A1)$ is satisfied, the user has to
For each $k$, to perform the rejection sampling, the user has to either
\begin{enumerate}
\item Find $\widehat{\sigma}_+^k$ by numerical optimization in the (A1) and (A2) case, and set for every particle $i$, $\widehat{\sigma}_+^{k, i } = \widehat{\sigma}_+^k$.
\item Or, for each particle $i$, find a $\widehat{\sigma}_+^{k, i } = \mathrm{sup}_{j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)$, in the (A3) case.
\end{enumerate}
For each $i$, the corresponding bound will be used to perform Lemma 1.\\
The first bound is uniform and do not depend on the particles sampled, therefore can be used for every particle $i$.  However, the resulting bound  can be "large"(with regards to the simulated set of particles) for the AR algorithm of Lemma 1.
The second bound requires $N$ computations per particle (therefore $N^2$ computations). However, it is clear that this second bound gives the optimal bound for Lemma 1.\\
Actually, the first bound is more likely to be "too large" when the time step is large. Furthermore, in that case, a rejection in Lemma 1 is more costly as it would require sampling another Brownian bridge skeleton (which would be large due to the large time step).\\
Overall, it seems difficult to have analytical result about which $\widehat{\sigma}_+^k$ to choose, as it depends both on the number of particles and the sampling time step, and that the bound given by (A3) is a random variable.\\
However, we did an empirical study on the SINE model, that satisfies A1 and A3. Results are presented in Figure \ref{Figure}.

 As expected from what is said above, for a large time step, even for a large number of particles, the bound given by A3, despite its quadratic cost, leads to less computationnal time than the uniform bound A1. 
 On the other hand, for a small time step, the uniform bound of A1 is good enough, and it's better to choose this bound instead of the optimal bound A3.
\begin{figure}[H]
\centering
\begin{tabular}{cc}
$\Delta = 0.5$ & $\Delta = 2$\\
\includegraphics[width = 0.4\textwidth]{Times_Delta05}&
\includegraphics[width = 0.4\textwidth]{Times_Delta2}
\end{tabular}
\caption{Empirical distribution of required time (in logarithmic scale of milliseconds) for the acceptation of a candidate in the acceptance rejection algorithm of Lemma 1, depending on which bound is chosen. A1 is when the uniform bound is chosen, (see the paper), A3 is when the particle based bound is chosen. One can see that the best choice depends on the number of particles (as expected) but also on the time step. A3 gives the optimal bound, but a larger computationnal cost.}\label{Figure}
\end{figure}
This is now mentionned (without this empirical study) in the numerical experiments section (in the SINE model subsection).  We believe that it might be confusing to insist too much on this point by adding this study.

@@

This was changed as suggested.

\item {\em page 11, line 10: does the fixed lag algorithm have the same cost for all lag values?
What is the computational time of one run of GRand PaRIS for the examples presented?}

\vspace{.3cm}

The only cost difference for two lags $\delta_1 < \delta_2$ is the memory space to keep tracks of particles between times $t+\delta_1$ and $t+\delta_2$ (basically, a $N*(\delta_2 - \delta_1)$ matrix). Therefore the computationnal time is identical for two different lags.\\
On a personnal laptop  (i7-6600U CPU @ 2.60GHz), for the parameters given in the paper, it took around 25 seconds for each E step. As mentionned in the paper, all the comparisons are made for a same amount of computation time (that is why there is more particles used in the fixed lag techniques). 
This is now clarified in the paper.

### 2017-09-14 TexSubmission/ParisEM.tex

\section{Conclusions}
This paper presents a new online SMC smoother for partially observed differential equations.
 This algorithm relies on an acceptance-rejection procedure inspired from the recent PaRIS algorithm.
 The main result of the article for practical applications is that the mechanism of this procedure remains valid when the transition density is approximated by a an unbiased positive estimator.
  The proposed procedure therefore extends the PaRIS algorithm to HMMs whose transition density is unknown and can be unbiasedly approximated. The GRand PaRIS algorithm outperforms the existing fixed lag smoother for POD processes of \cite{olsson:strojby:2011}, as it does not introduce any intrinsic and non vanishing bias.
   In addition, numerical simulations highlight a better variance using data from two different models.
   It can be implemented for the class of models $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} with a linear complexity in $N$.

>>>>

\section{Conclusions}
This paper presents a new online SMC smoother for partially observed differential equations.
 This algorithm relies on an acceptance-rejection procedure inspired from the recent PaRIS algorithm.
 The main result of the article for practical applications is that the mechanism of this procedure remains valid when the transition density is approximated by a an unbiased positive estimator.
  The proposed procedure therefore extends the PaRIS algorithm to HMMs whose transition density is unknown and can be unbiasedly approximated. The GRand PaRIS algorithm outperforms the existing fixed lag smoother for POD processes of \cite{olsson:strojby:2011}, as it does not introduce any intrinsic and non vanishing bias.
   In addition, numerical simulations highlight a better variance using data from two different models.
   It can be implemented for the class of models for which exact algorithms of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} are valids, with a linear complexity in $N$ in the best cases, or at worse in $N^2$.
